---
title: "Data 607 Final Project Analysis"
author: "Mohamed Hassan-El Serafi and Eddie Xu"
date: "2023-05-11"
output: html_document
---

# Introduction

The pressure of winning games can affect the performances of players and their respective teams. For this analysis, we will focus on the performance of players and teams during clutch situations, exploring how their performances are affected by pressure situations. More specifically, do teams/players perform better or worse from the regular season to the playoffs? 

The term "clutch" is defined as a situation when a game is within 5 points in the final 5 minutes. We will examine data that compares team clutch stats with non-clutch stats, focusing specifically on the regular season. Additionally, we will also analyze regular season and playoff clutch performances for both teams and players, respectively. The data we collected centered on 20 NBA seasons for team statistics from 2001-2002 to 2021-2022, and 10 NBA seasons for players from 2011-2012 to 2021-2022. We will use traditional statistics such as field goal and free throw percentages, win percentage, total points, and plus/minus, which measures whether the score margin increased or decreased with the team or player on the court. While there are numerous amounts of advanced stats, we will use only two metrics for this analysis: true shooting percentage, which uses field goal and free throw attempts to calculate a team and player's shooting percentages, and turnover rate, which is calculated utilizing turnovers, field goal and free throw attempts. This data was obtained using the hoopR and nbastatR libraries. 



# Loading Packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nbastatR)
library(hoopR)
library(ggfortify)
library(DT)
library(caret)
```




```{r}
new_team_csv <- read.csv("https://raw.githubusercontent.com/moham6839/Data607_Final_Project/master/Team_Clutch_Data.csv")
DT::datatable(new_team_csv)
```


```{r}
# split the dataset into regular season and playoff seasons
regular_season_dataset <- new_team_csv %>%
  filter(Season_Type == "Regular Season")

playoff_season_dataset <- new_team_csv %>%
  filter(Season_Type == "Playoffs")
```




# Team Clutch Stats - Regular Season 2001-02 to 2021-22







```{r}
team_rs_clutch <- regular_season_dataset %>%
  select(TEAM_NAME, PLUS_MINUS, FGM, FGA, FG3M, FG3A, FTM, FTA, GP, W, PTS, REB, TOV) %>%
  group_by(TEAM_NAME) %>%
  summarise(total_pm = sum(PLUS_MINUS),
  points_per_game = sum(PTS)/sum(GP),
  total_games = sum(GP),
  total_wins = sum(W),
  win_pct = sum(W)/sum(GP),
  total_points = sum(PTS),
  total_rebounds = sum(REB),
  total_fgm = sum(FGM),
  total_fga = sum(FGA),
  total_fgpct = sum(FGM)/sum(FGA),
  total_fg3pct = sum(`FG3M`)/sum(`FG3A`),
  total_ftm = sum(FTM),
  total_fta = sum(FTA),
  total_tov = sum(TOV),
  ft_pct = sum(FTM)/sum(FTA),
  #True shooting percentage: PT/(2*(FGA+0.44*FTA))
  true_shoot_pct = total_points/(2*(total_fga+0.44*total_fta)),
  # Turnover Rate = 100*TO/(FGA+0.44*FTA+TO)
  tov_rate = 100*total_tov /(total_fga+0.44*total_fta+total_tov)
  ) %>%
  arrange(desc(total_pm), win_pct)
DT::datatable(team_rs_clutch)
```




## Team Regular Season Clutch Winning Percentage


```{r}
ggplot(data=team_rs_clutch, aes(x=reorder(TEAM_NAME, -win_pct), y=win_pct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Regular Season Team Clutch Win Percentage",
       y="Win Percentage",
       x="Team") +
  coord_flip()
```



## Team Regular Season Clutch Total Plus/Minus



```{r}
ggplot(data=team_rs_clutch, aes(x=reorder(TEAM_NAME, -total_pm), y=total_pm)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Regular Season Team Clutch Plus/Minus",
       y="Plus/Minus",
       x="Team") +
  coord_flip()
```


## Team Regular Season Clutch Total Points Scored



```{r}
ggplot(data=team_rs_clutch, aes(x=reorder(TEAM_NAME, -total_points), y=total_points)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Regular Season Total Team Clutch Points",
       y="Total Points",
       x="Team") +
  coord_flip()
```



## Team Regular Season Clutch Field Goal Percentage


```{r}
ggplot(data=team_rs_clutch, aes(x=reorder(TEAM_NAME, -total_fgpct), y=total_fgpct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Regular Season Total Clutch Team Field Goal %",
       y="Field Goal Percentage",
       x="Team") +
  coord_flip()
```


## Team Regular Season Clutch Free Throw Percentage



```{r}
ggplot(data=team_rs_clutch, aes(x=reorder(TEAM_NAME, -ft_pct), y=ft_pct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Regular Season Team Clutch Free Throw %",
       y="Free Throw %",
       x="Team") +
  coord_flip()
```



## Team Regular Season Clutch True Shooting Percentage



```{r}
ggplot(data=team_rs_clutch, aes(x=reorder(TEAM_NAME, -true_shoot_pct), y=true_shoot_pct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Regular Season Total Clutch Team True Shooting %",
       y="True Shooting Percentage",
       x="Team") +
  coord_flip()
```



## Team Regular Season Clutch Turnover Rate



```{r}
ggplot(data=team_rs_clutch, aes(x=reorder(TEAM_NAME, -tov_rate), y=tov_rate)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Regular Season Total Clutch Turnover Rate",
       y="Turnover Rate",
       x="Team") +
  coord_flip()
```


## Team Regular Season Clutch Multiple Linear Regression Model


```{r}
rs_team_lm <- lm(win_pct ~ total_pm + true_shoot_pct + ft_pct + total_fgpct + tov_rate + total_points, data=team_rs_clutch)
```


```{r}
summary(rs_team_lm)
```


```{r}
rs_team_lm2 <- lm(win_pct ~ total_pm, data=team_rs_clutch)
```


```{r}
summary(rs_team_lm2)
```

















# Team Clutch Stats - Playoffs 2001-02 to 2021-22






```{r}
team_playoff_clutch <- playoff_season_dataset %>%
  select(TEAM_NAME, PLUS_MINUS, FGM, FGA, FG3M, FG3A, FTM, FTA, GP, W, PTS, REB, TOV) %>%
  group_by(TEAM_NAME) %>%
  summarise(total_pm = sum(PLUS_MINUS),
  points_per_game = sum(PTS)/sum(GP),
  total_games = sum(GP),
  total_wins = sum(W),
  win_pct = sum(W)/sum(GP),
  total_points = sum(PTS),
  total_rebounds = sum(REB),
  total_fgm = sum(FGM),
  total_fga = sum(FGA),
  total_fgpct = sum(FGM)/sum(FGA),
  total_fg3pct = sum(`FG3M`)/sum(`FG3A`),
  total_ftm = sum(FTM),
  total_fta = sum(FTA),
  total_tov = sum(TOV),
  ft_pct = sum(FTM)/sum(FTA),
  #True shooting percentage: PT/(2*(FGA+0.44*FTA))
  true_shoot_pct = total_points/(2*(total_fga+0.44*total_fta)),
  # Turnover Rate = 100*TO/(FGA+0.44*FTA+TO)
  tov_rate = 100*total_tov /(total_fga+0.44*total_fta+total_tov)
  ) %>%
  # using filter to remove outlier
  filter(win_pct < 0.8) %>%
  arrange(desc(total_pm), win_pct)
DT::datatable(team_playoff_clutch)
```





## Team Playoffs Clutch Winning Percentage




```{r}
ggplot(data=team_playoff_clutch, aes(x=reorder(TEAM_NAME, -win_pct), y=win_pct)) +
  geom_bar(stat="identity", position="dodge") +theme_minimal() +
  labs(title="NBA Playoffs Team Clutch Win Percentage",
       y="Win Percentage",
       x="Team") +
  coord_flip()
```


## Team Playoffs Clutch Total Plus/Minus



```{r}
ggplot(data=team_playoff_clutch, aes(x=reorder(TEAM_NAME, -total_pm), y=total_pm)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Team Clutch Plus/Minus",
       y="Plus/Minus",
       x="Team") +
  coord_flip()
```


## Team Playoffs Clutch Field Goal Percentage


```{r}
ggplot(data=team_playoff_clutch, aes(x=reorder(TEAM_NAME, -total_fgpct), y=total_fgpct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Team Clutch Field Goal %",
       y="Field Goal %",
       x="Team") +
  coord_flip()
```



## Team Playoffs Clutch Free Throw Percentage



```{r}
ggplot(data=team_playoff_clutch, aes(x=reorder(TEAM_NAME, -ft_pct), y=ft_pct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Team Clutch Free Throw %",
       y="Free Throw %",
       x="Team") +
  coord_flip()
```



## Team Playoffs Clutch Total Points


```{r}
ggplot(data=team_playoff_clutch, aes(x=reorder(TEAM_NAME, -total_points), y=total_points)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Team Clutch Total Points ",
       y="Total Points",
       x="Team") +
  coord_flip()
```


## Team Playoffs Clutch True Shooting Percentage


```{r}
ggplot(data=team_playoff_clutch, aes(x=reorder(TEAM_NAME, -true_shoot_pct), y=true_shoot_pct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Team Clutch True Shooting % ",
       y="True Shooting %",
       x="Team") +
  coord_flip()
```



## Team Playoffs Clutch Turnover Rate




```{r}
ggplot(data=team_playoff_clutch, aes(x=reorder(TEAM_NAME, -tov_rate), y=tov_rate)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Team Clutch Turnover Rate",
       y="Turnover Rate",
       x="Team") +
  coord_flip()
```



## Multiple Linear Regression Model - Team Playoff Clutch



```{r}
playoff_team_lm <- lm(win_pct ~ total_pm + total_fgpct+ ft_pct + true_shoot_pct + tov_rate + total_points, data=team_playoff_clutch)
```


```{r}
summary(playoff_team_lm)
```


```{r}
playoff_team_lm2 <- lm(win_pct ~ total_pm, data=team_playoff_clutch)
```


```{r}
summary(playoff_team_lm2)
```






# Player Clutch Data - Regular Season 2011-12 to 2021-22




```{r}
new_player_csv <- read.csv("https://raw.githubusercontent.com/moham6839/Data607_Final_Project/master/Player_Clutch_Data.csv")
head(new_player_csv)
```


```{r}
regular_season_player_dataset <- new_player_csv %>%
  filter(Season_Type == "Regular Season")
```








```{r}
rs_clutch_player <- regular_season_player_dataset %>%
  select(PLAYER_NAME, PLUS_MINUS, FGM, FGA, FG3M, FG3A, FTM, FTA, GP, W, PTS, REB, TOV) %>%
  group_by(PLAYER_NAME) %>%
  summarise(total_pm = sum(PLUS_MINUS),
  points_per_game = sum(PTS)/sum(GP),
  total_games = sum(GP),
  total_wins = sum(W),
  win_pct = sum(W)/sum(GP),
  total_points = sum(PTS),
  total_rebounds = sum(REB),
  total_fgm = sum(FGM),
  total_fga = sum(FGA),
  total_fgpct = sum(FGM)/sum(FGA),
  total_fg3pct = sum(`FG3M`)/sum(`FG3A`),
  total_ftm = sum(FTM),
  total_fta = sum(FTA),
  total_tov = sum(TOV),
  ft_pct = sum(FTM)/sum(FTA),
  #True shooting percentage: PT/(2*(FGA+0.44*FTA))
  true_shoot_pct = total_points/(2*(total_fga+0.44*total_fta)),
  # Turnover Rate = 100*TO/(FGA+0.44*FTA+TO)
  tov_rate = 100*total_tov /(total_fga+0.44*total_fta+total_tov)
  ) %>%
  arrange(desc(total_pm), win_pct)
DT::datatable(rs_clutch_player)
```






## Regular Season Top 25 Players Clutch Total Plus/Minus



```{r}
ggplot(data=rs_clutch_player[1:25, ], aes(x=reorder(PLAYER_NAME, -total_pm), y=total_pm)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Regular Season Top 25 Player Clutch Plus/Minus",
       y="Total Plus/Minus",
       x="Player") +
  coord_flip()
```




## Regular Season Top 25 Players Clutch Winning Percentage (Minimum 150 games)



```{r}
player_rs_win_pct <- rs_clutch_player %>%
  select(PLAYER_NAME, win_pct, total_games) %>%
  filter(total_games >= 150) %>%
  arrange(desc(win_pct))
DT::datatable(player_rs_win_pct)
```


```{r}
ggplot(data=player_rs_win_pct[1:25, ], aes(x=reorder(PLAYER_NAME, -win_pct), y=win_pct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Top 25 Regular Season Player Clutch Win %",
       y="Win Percentage",
       x="Player") +
  coord_flip()
```







```{r}
ggplot(data=rs_clutch_player[1:25, ], aes(x=reorder(PLAYER_NAME, -win_pct), y=win_pct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Top 25 Regular Season Player Clutch Win %",
       y="Win Percentage",
       x="Player") +
  coord_flip()
```



## Regular Season Top 25 Players Clutch Total Points


```{r}
player_rs_total_pts <- rs_clutch_player %>%
  select(PLAYER_NAME, total_points) %>%
  #filter(total_games >= 100) %>%
  arrange(desc(total_points))
DT::datatable(player_rs_total_pts)
```





```{r}
ggplot(data=player_rs_total_pts[1:25, ], aes(x=reorder(PLAYER_NAME, -total_points), y=total_points)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Regular Season Top 25 Players Clutch Points Per Game",
       y="Total Points",
       x="Player") +
  coord_flip()
```


## Regular Season Top 25 Players Clutch Total Points Per Game (Minimum 50 games)


```{r}
player_rs_ppg <- rs_clutch_player %>%
  select(PLAYER_NAME, points_per_game, total_games) %>%
  filter(total_games >= 50) %>%
  arrange(desc(points_per_game))
DT::datatable(player_rs_ppg)
```


```{r}
ggplot(data=player_rs_ppg[1:25, ], aes(x=reorder(PLAYER_NAME, -points_per_game), y=points_per_game)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Regular Season Top 25 Player Clutch Points Per Game",
       y="Points Per Game",
       x="Player") +
  coord_flip()
```







## Regular Season Top 25 Players Clutch Field Goal Percentage (Minimum 200 games and 200 Field Goal Attempts)



```{r}
player_rs_fgpct <- rs_clutch_player %>%
  select(PLAYER_NAME, total_fgpct, total_fga, total_games) %>%
  filter(total_games >= 200 & total_fga >=200) %>%
  arrange(desc(total_fgpct))
DT::datatable(player_rs_fgpct)
```



```{r}
ggplot(data=player_rs_fgpct[1:25, ], aes(x=reorder(PLAYER_NAME, -total_fgpct), y=total_fgpct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Regular Season Top 25 Player Clutch Field Goal %",
       y="Field Goal %",
       x="Player") +
  coord_flip()
```







## Regular Season Top 25 Players Clutch Free Throw Percentange (Minimum 200 games and 100 Free Throws)

```{r}
player_rs_ftpct <- rs_clutch_player %>%
  select(PLAYER_NAME, ft_pct, total_fta, total_games) %>%
  filter(total_games >= 200 & total_fta >=100) %>%
  arrange(desc(ft_pct))
DT::datatable(player_rs_ftpct)
```






```{r}
ggplot(data=player_rs_ftpct[1:25, ], aes(x=reorder(PLAYER_NAME, -ft_pct), y=ft_pct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Regular Season Top 20 Player Clutch Free Throw %",
       y="Free Throw %",
       x="Player") +
  coord_flip()
```


## Regular Season Top 25 Players Clutch True Shooting Percentage (Minimum 200 Field Goal Attempts and 100 Free Throw Attempts)


```{r}
player_rs_tspct <- rs_clutch_player %>%
  select(PLAYER_NAME, true_shoot_pct, total_fga, total_fta) %>%
  filter(total_fga >= 200 & total_fta >= 100) %>%
  arrange(desc(true_shoot_pct))
DT::datatable(player_rs_tspct)
```



```{r}
ggplot(data=player_rs_tspct[1:25, ], aes(x=reorder(PLAYER_NAME, -true_shoot_pct), y=true_shoot_pct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Regular Season Top 25 Player Clutch True Shooting %",
       y="True Shooting %",
       x="Player") +
  coord_flip()
```



## Regular Season Top 25 Players Clutch Turnover Rate



```{r}
ggplot(data=rs_clutch_player[1:25, ], aes(x=reorder(PLAYER_NAME, -tov_rate), y=tov_rate)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Regular Season Top 25 Players Clutch Turnover Rate",
       y="Turnover Rate",
       x="Player") +
  coord_flip()
```



## Multiple Linear Regression Model - Regular Season Player Clutch Stats


```{r}
rs_player_lm <- lm(win_pct ~ total_pm + total_fgpct + ft_pct + true_shoot_pct + tov_rate + points_per_game, data=rs_clutch_player)
```



```{r}
summary(rs_player_lm)
```

```{r}
rs_player_lm2 <- lm(win_pct ~ total_pm +  true_shoot_pct, data=rs_clutch_player)
```



```{r}
summary(rs_player_lm2)
```









## Player Clutch Data - Playoffs 2011-12 to 2021-22



```{r}
playoff_season_player_dataset <- new_player_csv %>%
  filter(Season_Type == "Playoffs")
```



```{r}
playoff_performers <- playoff_season_player_dataset %>%
  select(PLAYER_NAME, PLUS_MINUS, FGM, FGA, FG3M, FG3A, FTM, FTA, GP, W, PTS, REB, TOV) %>%
  group_by(PLAYER_NAME) %>%
  summarise(total_pm = sum(PLUS_MINUS),
  points_per_game = sum(PTS)/sum(GP),
  total_games = sum(GP),
  total_wins = sum(W),
  win_pct = sum(W)/sum(GP),
  total_points = sum(PTS),
  total_rebounds = sum(REB),
  total_fgm = sum(FGM),
  total_fga = sum(FGA),
  total_fgpct = sum(FGM)/sum(FGA),
  total_fg3pct = sum(`FG3M`)/sum(`FG3A`),
  total_ftm = sum(FTM),
  total_fta = sum(FTA),
  total_tov = sum(TOV),
  ft_pct = sum(FTM)/sum(FTA),
  #True shooting percentage: PT/(2*(FGA+0.44*FTA))
  true_shoot_pct = total_points/(2*(total_fga+0.44*total_fta)),
  # Turnover Rate = 100*TO/(FGA+0.44*FTA+TO)
  tov_rate = 100*total_tov /(total_fga+0.44*total_fta+total_tov)
  ) %>%
  arrange(desc(total_pm), win_pct)
DT::datatable(playoff_performers)
```



## Playoffs Top 20 Players Clutch Total Plus/Minus








```{r}
ggplot(data=playoff_performers[1:20, ], aes(x=reorder(PLAYER_NAME, -total_pm), y=total_pm)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Top 20 Players Clutch Plus/Minus",
       y="Plus/Minus",
       x="Player") +
  coord_flip()
```


## Playoffs Bottom 25 Players Clutch Total Plus/Minus




```{r}
ggplot(data=playoff_performers[494:514, ], aes(x=reorder(PLAYER_NAME, total_pm), y=total_pm)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Bottom 20 Players Clutch Plus/Minus",
       y="Plus/Minus",
       x="Player") +
  coord_flip()
```





## Playoffs Top 20 Players Clutch Winning Percentage (Minimum 20 Playoff Games)




```{r}
player_playoff_win_pct <- playoff_performers %>%
  select(PLAYER_NAME, win_pct, total_games) %>%
  filter(total_games >= 20) %>%
  arrange(desc(win_pct))
DT::datatable(player_win_pct)
```



```{r}
ggplot(data=player_playoff_win_pct[1:20, ], aes(x=reorder(PLAYER_NAME, -win_pct), y=win_pct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Top 20 Players Clutch Win %",
       y="Win %",
       x="Player") +
  coord_flip()
```






## Playoffs Bottom 20 Players Clutch Winning Percentage

```{r}
ggplot(data=player_playoff_win_pct[71:91, ], aes(x=reorder(PLAYER_NAME, win_pct), y=win_pct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Bottom 20 Players Clutch Win %",
       y="Win %",
       x="Player") +
  coord_flip()
```







## Playoffs Top 20 Players Clutch Field Goal Percentage (Minimum 20 games and 40 Field Goal Attempts)

```{r}
player_playoff_fg_pct <- playoff_performers %>%
  select(PLAYER_NAME, total_fgpct, total_games, total_fga) %>%
  filter(total_games >= 20 & total_fga >= 40) %>%
  arrange(desc(total_fgpct))
DT::datatable(player_fg_pct)
```




```{r}
ggplot(data=player_playoff_fg_pct[1:20, ], aes(x=reorder(PLAYER_NAME, -total_fgpct), y=total_fgpct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Top 20 Players Clutch Field Goal %",
       y="Win %",
       x="Player") +
  coord_flip()
```









## Playoffs Bottom 20 Players Clutch Field Goal Percentage




```{r}
ggplot(data=player_playoff_fg_pct[20:40, ], aes(x=reorder(PLAYER_NAME, total_fgpct), y=total_fgpct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Bottom 20 Players Clutch Field Goal %",
       y="Field Goal %",
       x="Player") +
  coord_flip()
```



## Playoffs Player Clutch Free Throw Percentage (Minimum 25 Free Throw Attempts)


```{r}
player_playoff_ft_pct <- playoff_performers %>%
  select(PLAYER_NAME, ft_pct, total_fta) %>%
  filter(total_fta >= 25) %>%
  arrange(desc(ft_pct))
DT::datatable(player_ft_pct)
```




```{r}
ggplot(data=player_ft_pct[1:32, ], aes(x=reorder(PLAYER_NAME, -ft_pct), y=ft_pct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title= "NBA Playoffs Players Clutch Free Throw %",
       y="Free Throw %",
       x="Player") +
  coord_flip()
```





## Playoffs Top 25 Players Clutch Total Points (Minimum 20 games)

```{r}
player_playoff_total_pts <- playoff_performers %>%
  select(PLAYER_NAME, total_points, total_games) %>%
  filter(total_games >= 20) %>%
  arrange(desc(total_points))
DT::datatable(player_playoff_total_pts)
```




```{r}
ggplot(data=player_playoff_total_pts[1:20, ], aes(x=reorder(PLAYER_NAME, -total_points), y=total_points)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Top 20 Players Clutch Points",
       y="Points",
       x="Player") +
  coord_flip()
```



## Playoffs Top 25 Players Clutch Points Per Game (Minimum 20 games)


```{r}
player_playoff_ppg <- playoff_performers %>%
  select(PLAYER_NAME, points_per_game, total_games) %>%
  filter(total_games >= 20) %>%
  arrange(desc(points_per_game))
DT::datatable(player_playoff_ppg)
```




```{r}
ggplot(data=player_playoff_ppg[1:20, ], aes(x=reorder(PLAYER_NAME, -points_per_game), y=points_per_game)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Top 20 Players Clutch Points Per Game",
       y="Points Per Game",
       x="Player") +
  coord_flip()
```


## Playoffs Bottom 20 Players Clutch Points Per Game

```{r}
ggplot(data=player_playoff_ppg[71:91, ], aes(x=reorder(PLAYER_NAME, points_per_game), y=points_per_game)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Bottom 20 Players Clutch Points Per Game",
       y="Points Per Game",
       x="Player") +
  coord_flip()
```




## Playoffs Top 20 Players Clutch True Shooting Percentage (Minimum 50 Field Goal Attempts and 25 Free Throw Attempts)

```{r}
player_playoff_ts_pct <- playoff_performers %>%
  select(PLAYER_NAME, true_shoot_pct, total_fga, total_fta) %>%
  filter(total_fga >= 50 & total_fta >= 25) %>%
  arrange(desc(true_shoot_pct))
DT::datatable(player_playoff_ts_pct)
```



```{r}
ggplot(data=player_playoff_ts_pct[1:20, ], aes(x=reorder(PLAYER_NAME, -true_shoot_pct), y=true_shoot_pct)) +
  geom_bar(stat="identity", position="dodge") +
  theme_minimal() +
  labs(title="NBA Playoffs Top Players True Shooting %",
       y="True Shooting %",
       x="Player") +
  coord_flip()
```








## Multiple Linear Regression of Playoff Player Clutch Stats



```{r}
( lr.out <- glm(`win_pct` ~ total_plus_minus, data = playoff_performers, family = binomial(link = 'logit')) )
```




```{r}
playoff_player_lm <- lm(win_pct ~ total_plus_minus + total_fgm  + total_ftm + total_rebounds + total_points, data=playoff_performers)
```









```{r}
summary(playoff_player_lm)
```


```{r}
ggplot(data=playoff_performers, aes(playoff_player_lm$residuals)) +
geom_histogram(binwidth = 1, color = "black", fill = "purple4") +
theme(panel.background = element_rect(fill = "white"),
axis.line.x=element_line(),
axis.line.y=element_line()) +
ggtitle("Histogram for Model Residuals")
```



```{r}
ggplot(data = playoff_performers, aes(x = total_plus_minus, y = total_wins)) +
geom_point() +
stat_smooth(method = "lm", col = "dodgerblue3") +
theme(panel.background = element_rect(fill = "white"),
axis.line.x=element_line(),
axis.line.y=element_line()) +
ggtitle("Linear Model Fitted to Data")
```



```{r}
model <- train(total_wins ~ total_plus_minus,
               data = playoff_performers,
               method = "lm")
```





```{r}
player_lm <- lm(`win_pct` ~ `total_plus_minus`, data=playoff_performers)
```



```{r}
summary(player_lm)
```



## Takeaways






## Next Steps


W







